### 2021-02-06 analysis/paper.Rmd
Add skeleton document for analysis

---
title: "openarchaeoCollaboration"
description: |
  Insert abstract here.
author:
  - name: Zack Batist
    url: https://zackbatist.github.io
    affiliation: University of Toronto
  - name: Joe Roe
    url: https://joeroe.io
    affiliation: University of Copenhagen
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

### 2021-02-06 analysis/paper.Rmd
Start analysis of GitHub API data

@@
 ```
+
+```{r dependencies, include=FALSE}
+library("tidyverse")
+library("openarchaeoCollaboration")
+```
+
+## Materials and methods
+
+We used *open-archaeo* <https://open-archaeo.info/>, a list of open source archaeological software and other digital resources.
+The criteria for inclusion in open-archaeo are [...].
+The search strategy for compiling it was [...].
+
+```{r data-oarch}
+oarch <- read_csv("https://raw.githubusercontent.com/zackbatist/open-archaeo/master/open-archaeo.csv")
+```
+
+We augmented this with data from the GitHub API for records with an associated GitHub repository.
+
+```{r data-github}
+oarch %>% 
+  mutate(
+    gh_repo = str_remove(github, "https://github.com/"),
+    gh_repo = str_remove(gh_repo, "/$")
+  ) %>% 
+  separate(gh_repo, c("gh_owner", "gh_repo"), "/", extra = "drop") %>% 
+  mutate(
+    gh_lang = ght("GET /repos/{owner}/{repo}/languages", 
+                  owner = gh_owner, repo = gh_repo)
+  ) ->
+  oarch
+```
+
+## Results
+
+What languages are people using?
+
+```{r plot-gh_lang}
+oarch$gh_lang %>% 
+  pivot_longer(everything(), names_to = "language", values_to = "n") %>% 
+  drop_na(n) %>% 
+  group_by(language) %>% 
+  summarise(n = sum(n), .groups = "drop") %>% 
+  arrange(-n) %>% 
+  slice(1:10) %>% 
+  ggplot(aes(fct_reorder(language, n, .desc = TRUE), n)) +
+  geom_col() +
+  labs(title = "Languages used in GitHub repositories",
+       caption = "Data: <https://open-archaeo.info> and the GitHub API",
+       x = "Language", y = "Bytes of code")
+```

### 2021-02-06 R/gh.R

@@ -1,0 +1,46 @@
+# gh.R
+# Vectorised wrapper functions for the GitHub API
+
+#' Query the GitHub API and return a table
+#'
+#' @description
+#' Vectorised wrapper for [gh::gh()] that:
+#'
+#' * Converts the response to a data frame
+#' * Empty responses are returned as an `NA` row.
+#' * Silently skips queries where any value of `...` is `NA`, returning an `NA`
+#'   row.
+#' * Catches API error statuses (e.g. 404), returning an `NA` row and a printed
+#'   message instead of an error.
+#'
+#' @param endpoint GitHub API endpoint.
+#' @param ... Other arguments to [gh::gh()], including name-value pairs of any
+#'  API parameters in `endpoint`.
+#'
+#' @return
+#' A data frame with one row per query, and one column per response vector.
+#'
+#' @export
+#'
+#' @examples
+#' ght("GET /repos/{owner}/{repo}/languages",
+#'     owner = c("joeroe", "joeroe"), repo = c("era", "c14"))
+ght <- function(endpoint, ...) {
+  df <- pmap_dfr(data.frame(...), function(endpoint, ...) {
+    if (any(is.na(list(...)))) data.frame(`NA` = NA)
+    else {
+      res <- tryCatch(
+        gh(endpoint, ...),
+        http_error_404 = function(c) {
+          print(c)
+          data.frame(`NA` = NA)
+        }
+      )
+      if (length(res) > 0) res
+      else data.frame(`NA` = NA)
+    }
+  }, endpoint = endpoint)
+
+  if ("NA" %in% names(df)) return(select(df, -`NA`))
+  else return(df)
+}

### 2021-02-25 R/gh.R
Allow ght() to return a list of tibbles
If the response contains more than one row.

@@ -1,46 +1,53 @@
 # gh.R
 # Vectorised wrapper functions for the GitHub API
 
-#' Query the GitHub API and return a table
+#' Query the GitHub API (tabulated)
 #'
-#' @description
-#' Vectorised wrapper for [gh::gh()] that:
-#'
-#' * Converts the response to a data frame
-#' * Empty responses are returned as an `NA` row.
-#' * Silently skips queries where any value of `...` is `NA`, returning an `NA`
-#'   row.
-#' * Catches API error statuses (e.g. 404), returning an `NA` row and a printed
-#'   message instead of an error.
+#' [dplyr::mutate()]-friendly wrapper for [gh::gh()] that tabulates the response
+#' and always returns a vector that matches the number of queries.
+#' Vectorised over the named API parameters in `...`.
 #'
 #' @param endpoint GitHub API endpoint.
 #' @param ... Other arguments to [gh::gh()], including name-value pairs of any
 #'  API parameters in `endpoint`.
 #'
+#' @details
+#' Empty responses return an `NA` row, with a warning if the request produced
+#' a 404 error. Queries where any value of `...` is `NA` will also return an
+#' `NA` row.
+#'
 #' @return
-#' A data frame with one row per query, and one column per response vector.
+#' If each response returns a single row, a `tibble` with one row per query
+#' (producing a packed column if used in `mutate()`). Otherwise, a list of
+#' `tibbles` for each query (producing a nested column if used in `mutate()`).
 #'
 #' @export
 #'
 #' @examples
-#' ght("GET /repos/{owner}/{repo}/languages",
-#'     owner = c("joeroe", "joeroe"), repo = c("era", "c14"))
+#' ght("GET /repos/{owner}/{repo}/languages", owner = c("joeroe", "joeroe"), repo = c("era", "c14"))
 ght <- function(endpoint, ...) {
-  df <- pmap_dfr(data.frame(...), function(endpoint, ...) {
+  responses <- purrr::pmap(data.frame(...), function(endpoint, ...) {
     if (any(is.na(list(...)))) data.frame(`NA` = NA)
     else {
       res <- tryCatch(
         gh(endpoint, ...),
-        http_error_404 = function(c) {
-          print(c)
-          data.frame(`NA` = NA)
-        }
+        http_error_404 = gh_warn_404
       )
-      if (length(res) > 0) res
[…]

### 2021-02-25 analysis/paper.Rmd
Initial analysis of contributors

@@
 ```{r data-github}
+# Pretty slow
 oarch %>% 
@@
   ) %>% 
-  separate(gh_repo, c("gh_owner", "gh_repo"), "/", extra = "drop") %>% 
+  separate(gh_repo, c("gh_owner", "gh_repo"), "/", 
+           fill = "right", extra = "drop") %>% 
   mutate(
     gh_lang = ght("GET /repos/{owner}/{repo}/languages", 
+                  owner = gh_owner, repo = gh_repo),
+    gh_contributors = ght("GET /repos/{owner}/{repo}/contributors",
                           owner = gh_owner, repo = gh_repo)
@@
 
-What languages are people using?
+### What languages are people using?
 
@@
 ```
+
+### Who is contributing?
+
+```{r plot-gh_contributors-by-repo}
+oarch %>% 
+  unnest(gh_contributors) %>% 
+  mutate(gh_repo = paste(gh_owner, gh_repo, sep = "/")) %>% 
+  group_by(gh_repo) %>% 
+  summarise(n = n(), .groups = "drop") %>% 
+  ggplot(aes(fct_reorder(gh_repo, n, .desc = TRUE), n)) +
+  geom_col() +
+  labs(
+    title = "Repositories by number of contributors",
+    caption = "Data: <https://open-archaeo.info> and the GitHub API",
+    x = "Repository", y = "Number of contributors"
+  ) +
+  theme(axis.text.x = element_blank())
+```

### 2021-02-25 analysis/paper.Rmd
Wrote initial draft for an abstract.

@@
 ```
+
+## Abstract
+Computational tools and practices are increasingly recognized as valuable and commonplace components of the general archaeological toolbox.
+Aside from writing one-off scripts used to process and analyze data pertaining to specific research programs, archaeologists have also been developing software designed for general use.
+This paper surveys the field of archaeological software development in order to understand what kinds of software archaeologists are making, and to identify norms regarding how software is built and maintained.
+Specifically, we focus our inquiry on collaborative aspects of software development, and how archaeologists make use of tools and platforms designed to facilitate collaboration among professional software developers.
+Our formal analysis of specific engagements with software development projects hosted on GitHub finds that [...]
 

### 2021-02-27 R/gh.R
Replace ght() with specific functions for each GitHub API endpoint
Scrapping the generalised approach; the responses are too variable.

@@ -1,53 +1,178 @@
 # gh.R
-# Vectorised wrapper functions for the GitHub API
+# Functions for accessing the GitHub API
 
-#' Query the GitHub API (tabulated)
+#' GitHub repository queries
 #'
-#' [dplyr::mutate()]-friendly wrapper for [gh::gh()] that tabulates the response
-#' and always returns a vector that matches the number of queries.
-#' Vectorised over the named API parameters in `...`.
+#' Get the languages, contributors, issues, comments, or commits of a single
+#' repository from the GitHub API.
+#'
+#' @param repo Name of the repository, in the form `"owner/repository"`.
+#'
+#' @return
+#' A `tibble` containing the tabulated API responses, or `NA` if the response
+#' was empty or unsuccessful.
+#'
+#' @details
+#' Queries use [ghq()], a thin wrapper for [gh::gh()] that ensures the query
+#' always returns a value.
+#'
+#' @name gh_query
+#'
+#' @examples
+#' gh_lang("joeroe/era")
+#' gh_contrib("joeroe/era")
+#' gh_issue("joeroe/era")
+#' gh_comment("joeroe/era")
+#' gh_commit("joeroe/era")
+NULL
+
+#' @rdname gh_query
+#' @export
+gh_lang <- function(repo) {
+  data <- tibble::tibble(
+    repo = repo,
+    response = ghq("GET /repos/{repo}/languages", repo = repo)
+  )
+
+  if (not_all_na(data$response)) {
+    data <- dplyr::mutate(data, lang = names(response))
+    data <- tidyr::unnest(data, response)
+    data <- dplyr::select(data, repo, lang, bytes = response)
+    data <- dplyr::select(data, where(not_all_na))
+    data
+  }
+  else NA
+}
+
+#' @rdname gh_query
+#' @export
+gh_contrib <- function(repo) {
+  data <- tibble::tibble(
+    repo = repo,
+    response = ghq("GET /repos/{repo}/contributors", repo = repo)
+  )
+
+  if (not_all_na(data$response)) {
+    data <- tidyr::hoist(data, response,
+                         contributor = "login",
+                         contributions = "contributions")
+    data <- dplyr::select(data, -response)
+    data
+  }
+  else NA
+}
+
+#' @rdname gh_query
+#' @export
+gh_issue <- function(repo) {
+  data <- tibble::tibble(
+    repo = repo,
+    response = ghq("GET /repos/{repo}/issues", repo = repo)
+  )
+
[…]

### 2021-02-27 analysis/paper.Rmd
Update GitHub API data
And backup intermediate data in analysis/derived_data/

@@
 ```{r data-oarch}
-oarch <- read_csv("https://raw.githubusercontent.com/zackbatist/open-archaeo/master/open-archaeo.csv")
+read_csv("https://raw.githubusercontent.com/zackbatist/open-archaeo/master/open-archaeo.csv") %>% 
+  mutate(category = na_if(category, "?")) %>% 
+  mutate(gh_repo = str_remove(github, "https://github.com/")) ->
+  oarch
 ```
@@
 ```{r data-github}
-# Pretty slow
-oarch %>% 
-  mutate(
-    gh_repo = str_remove(github, "https://github.com/"),
-    gh_repo = str_remove(gh_repo, "/$")
-  ) %>% 
-  separate(gh_repo, c("gh_owner", "gh_repo"), "/", 
-           fill = "right", extra = "drop") %>% 
-  mutate(
-    gh_lang = ght("GET /repos/{owner}/{repo}/languages", 
-                  owner = gh_owner, repo = gh_repo),
-    gh_contributors = ght("GET /repos/{owner}/{repo}/contributors",
-                          owner = gh_owner, repo = gh_repo)
-  ) ->
-  oarch
+# Warning: slow
+# TODO: map
+gh_langs <- map(oarch$gh_repo, gh_lang)
+gh_contribs <- map(oarch$gh_repo, gh_contrib)
+gh_issues <- map(oarch$gh_repo, gh_issue)
+gh_comments <- map(oarch$gh_repo, gh_comment)
+gh_commits <- map(oarch$gh_repo, function(x) { print(x); gh_commit(x) })
 ```
@@
 
-### What languages are people using?
+### Tags and categories
+
+```{r plot-categories}
+oarch %>% 
+  group_by(category) %>% 
+  drop_na(category) %>% 
+  summarise(n = n()) %>% 
+  ggplot(aes(fct_reorder(category, n), n)) +
+  geom_col() +
+  coord_flip() +
+  labs(
+    title = "Open archaeology categories",
+    caption = "Data: <https://open-archaeo.info>",
+    x = NULL, y = "Number of projects"
+  )
+```
+
+```{r plot-tags}
+oarch %>% 
+  pivot_longer(c(tag1, tag2, tag3, tag4, tag5), values_to = "tag", 
+               values_drop_na = TRUE) %>% 
+  group_by(tag) %>%
+  drop_na(tag) %>%
+  summarise(n = n()) %>%
+  ggplot(aes(fct_reorder(tag, n), n)) +
+  geom_col() +
+  coord_flip() +
+  labs(
+    title = "Open archaeology tags",
+    caption = "Data: <https://open-archaeo.info>",
+    x = NULL, y = "Number of projects"
+  )
+```
+
+### Use of GitHub
 
@@
 
-### Who is contributing?
+### Repository contributors
 
-```{r plot-gh_contributors-by-repo}
+```{r plot-gh_contributors-histogram}
 oarch %>% 
+  drop_na(gh_repo) %>% 
   unnest(gh_contributors) %>% 
@@
   summarise(n = n(), .groups = "drop") %>% 
-  ggplot(aes(fct_reorder(gh_repo, n, .desc = TRUE), n)) +
-  geom_col() +
+  ggplot(aes(x = n)) +
+  geom_histogram(binwidth = 1) +
   labs(
@@
     caption = "Data: <https://open-archaeo.info> and the GitHub API",
-    x = "Repository", y = "Number of contributors"
-  ) +
-  theme(axis.text.x = element_blank())
+    x = "Contributors", y = "Repositories"
+  )
 ```

### 2021-02-27 analysis/paper.Rmd
Keep GitHub data in main table

@@
 # TODO: map
-gh_langs <- map(oarch$gh_repo, gh_lang)
-gh_contribs <- map(oarch$gh_repo, gh_contrib)
-gh_issues <- map(oarch$gh_repo, gh_issue)
-gh_comments <- map(oarch$gh_repo, gh_comment)
-gh_commits <- map(oarch$gh_repo, function(x) { print(x); gh_commit(x) })
+oarch <- mutate(oarch,
+                gh_langs = map(oarch$gh_repo, gh_lang),
+                gh_contribs = map(oarch$gh_repo, gh_contrib),
+                gh_issues = map(oarch$gh_repo, gh_issue),
+                gh_comments = map(oarch$gh_repo, gh_comment),
+                gh_commits = map(oarch$gh_repo, gh_commit),
+)
 ```

### 2021-02-27 analysis/paper.Rmd
Update plots for new GitHub data format

@@
 
-```{r plot-gh_lang}
-oarch$gh_lang %>% 
-  pivot_longer(everything(), names_to = "language", values_to = "n") %>% 
-  drop_na(n) %>% 
-  group_by(language) %>% 
-  summarise(n = sum(n), .groups = "drop") %>% 
+```{r plot-gh_langs, fig.show="hold", out.width="50%"}
+oarch %>% 
+  unnest(gh_langs) %>% 
+  drop_na(lang, bytes) %>% 
+  group_by(lang) %>% 
+  summarise(n = n(), .groups = "drop") %>% 
   arrange(-n) %>% 
   slice(1:10) %>% 
-  ggplot(aes(fct_reorder(language, n, .desc = TRUE), n)) +
+  ggplot(aes(fct_reorder(lang, n, .desc = TRUE), n)) +
   geom_col() +
   labs(title = "Languages used in GitHub repositories",
+       subtitle = "Number of repositories using top 10 languages",
        caption = "Data: <https://open-archaeo.info> and the GitHub API",
-       x = "Language", y = "Bytes of code")
+       x = NULL, y = "Number of repositories")
+
+oarch %>% 
+  unnest(gh_langs) %>% 
+  drop_na(lang, bytes) %>% 
+  group_by(lang) %>% 
+  summarise(bytes = sum(bytes), .groups = "drop") %>% 
+  arrange(-bytes) %>% 
+  slice(1:10) %>% 
+  ggplot(aes(fct_reorder(lang, bytes, .desc = TRUE), bytes)) +
+  geom_col() +
+  labs(title = "Languages used in GitHub repositories",
+       subtitle = "Total bytes of code in top 10 languages",
+       caption = "Data: <https://open-archaeo.info> and the GitHub API",
+       x = NULL, y = "Bytes")
 ```
@@
   drop_na(gh_repo) %>% 
-  unnest(gh_contributors) %>% 
-  mutate(gh_repo = paste(gh_owner, gh_repo, sep = "/")) %>% 
+  unnest(gh_contribs) %>% 
   group_by(gh_repo) %>% 
@@
   labs(
-    title = "Repositories by number of contributors",
+    title = "Repository contributors",
     caption = "Data: <https://open-archaeo.info> and the GitHub API",
-    x = "Contributors", y = "Repositories"
+    x = "Number of contributors", y = "Number of repositories"
   )

### 2021-02-27 analysis/paper.Rmd
Add plot of commits through time
And reknit paper.Rmd

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
+                      layout = "l-body-outset")
 ```
@@
 library("tidyverse")
+library("lubridate")
 library("openarchaeoCollaboration")
@@
   oarch
+
+oarch <- mutate(oarch, rough_category = recode(category,
+                                 "R" = "Packages (R, Python, etc.)",
+                                 "Web apps" = "Standalone apps",
+                                 "Datasets" = "Analyses & datasets",
+                                 "Tutorials" = "Lists & tutorials",
+                                 "Python" = "Packages (R, Python, etc.)",
+                                 "Desktop apps" = "Standalone apps",
+                                 "QGIS" = "Plugins",
+                                 "Lists" = "Lists & tutorials",
+                                 "Mobile apps" = "Standalone apps",
+                                 "Analyses" = "Analyses & datasets",
+                                 "Spreadsheets" = "Forms & spreadsheets",
+                                 "Command line tools" = "Standalone apps",
+                                 "Plugins" = "Plugins",
+                                 "ArcGis" = "Plugins",
+                                 "Filemaker" = "Plugins",
+                                 "ODK" = "Forms & spreadsheets",
+                                 "Lisp" = "Packages (R, Python, etc.)",
+                                 .default = "Other",
+                                 .missing = "Other"
+                                 ))
 ```
@@
 
+```{r plot-gh_commits-cum}
+oarch %>% 
+  unnest(gh_commits) %>% 
+  drop_na(sha) %>% 
+  mutate(date = floor_date(datetime, "week")) %>% 
+  mutate(rough_category = recode(
+    rough_category,
+    "Standalone apps" = "Standalone apps",
+    "Packages (R, Python, etc.)" = "Packages (R, Python, etc.)",
+    .default = "Other")) %>% 
+  group_by(rough_category, date) %>% 
+  summarise(n_commits = n(), .groups = "drop_last") %>% 
+  mutate(cat_total = sum(n_commits)) %>% 
+  mutate(cum_commits = cumsum(n_commits)) %>% 
+  ggplot(aes(date, cum_commits, 
+             fill = fct_reorder(rough_category, cat_total))) +
+  geom_area() +
+  scale_x_datetime(limits = c(ymd("2008-01-01", tz = "UTC"), NA),
+                   date_breaks = "2 years",
+                   date_labels = "%Y") +
+  scale_fill_discrete(guide = guide_legend(reverse = TRUE)) +
+  labs(
+    title = "Open source archaeology",
+    subtitle = "Cumulative commits in GitHub repositories",
+    caption = "Data: <https://open-archaeo.info> and the GitHub API",
+    x = NULL, y = "Commits", fill = NULL
+  ) +
+  theme(legend.position = "bottom")
+```
+
 ```{r plot-gh_langs, fig.show="hold", out.width="50%"}

### 2021-03-01 analysis/paper.Rmd
Some progress towards the abstract.

@@
 description: |
-  Insert abstract here.
+  Advanced digital practices involving the production and manipulation of computer code are gradually becoming ubiquitous aspects of archaeological research.
+The relative simplicity afforded by scientific scripting languages such as R and Python, which enable custom code to be written, modified, commented, shared and reused with ease, has been a boon for fostering greater digital data manipulation skills among archaeologists.
+Code used to support archaeological research largely comprise custom scripts or project files that pass data through functions that parse and transform inputs according to documented routines.
+Base functions or functions included in generic data manipulation, analysis and visualization packages (such as those that collectively form the Tidyverse, ROpenSci and [insert Python equivalent] ecosystems, or 'off the shelf' database management systems or visualization engines) typically support a broad set of common applications, however these tools are rarely designed with archaeological use cases in mind.
+Recognizing the need for specialized software that more effectively suits archaeological needs, archaeologists have taken it upon themselves to create tools specifically designed for archaeological use cases.
+These tools operate by abstracting the factors and desired outcomes that the designers perceive to be commonly held by intended users, which happen to be archaeologists.
+As series of generic functions that can operate with any appropriate input data, this class of software differentiates itself from custom one-off scripts used to process and analyze data pertaining to specific research projects.
+
+  NOTE: Refer to https://github.com/benmarwick/ctv-archaeology for an overview of common archaeological use cases, and which highlights how generic and archaeology-specific functions may be selected and applied for various applications.
+This is specific to R, but refer also to blogs, web resources and twitter conversations in general that discuss either how to accomplish archaeology-specific tasks using generic toolsets.
+Basically demonstrate that the gaps that archaeologists identify are wrapped up in and perhaps even derived from conversations about their experiences writing and manipulating code.
+
+  This paper surveys the field of archaeological software development in order to understand what kinds of general use digital tools archaeologists are making and to identify norms regarding how software is built and maintained -- focusing specifically on collaborative aspects of software development.
+We use open-archaeo, a curated list of open source archaeological software and other digital resources, as the basis of this survey.
+This dataset is augmented with data collected from software development productivity platforms such as GitHub, which are used to host and facilitate collaborative software development work.
+Our analysis examines trends relating to the frequency of documented work, the distribution of contributions made by collaborating participants, how software are documented, packaged and presented, and the use of tools designed to support collaborative workflows.
+
+  This work complements other studies that investigate scholarly use of software development productivity platforms in their work that rely on focus group sessions, questionnaires and interviews (such as [IASGE](https://investigating-archiving-git.gitlab.io/)).
+
 author:
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
-                      layout = "l-body-outset")
+knitr::opts_chunk$set(
+	echo = FALSE,
+	message = FALSE,
+	warning = FALSE,
+	layout = "l-body-outset"
+)
 ```
@@
 ```
-
-## Abstract
-Computational tools and practices are increasingly recognized as valuable and commonplace components of the general archaeological toolbox.
-Aside from writing one-off scripts used to process and analyze data pertaining to specific research programs, archaeologists have also been developing software designed for general use.
-This paper surveys the field of archaeological software development in order to understand what kinds of software archaeologists are making, and to identify norms regarding how software is built and maintained.
-Specifically, we focus our inquiry on collaborative aspects of software development, and how archaeologists make use of tools and platforms designed to facilitate collaboration among professional software developers.
-Our formal analysis of specific engagements with software development projects hosted on GitHub finds that [...]
 

### 2021-03-03 analysis/paper.Rmd
Move CAA abstract to own file
Since it's long and also shouldn't have our names on it for review.

--- 2021-02-27 analysis/paper.Rmd
@@
 title: "openarchaeoCollaboration"
-description: |
-  Insert abstract here.
 author:

### 2021-03-12 analysis/paper.Rmd
Refine gh_commits-cum plot aesthetics

@@
 library("openarchaeoCollaboration")
+library("khroma")
 ```
@@
 # Warning: slow
-# TODO: map
 oarch <- mutate(oarch,
@@
                 gh_comments = map(oarch$gh_repo, gh_comment),
-                gh_commits = map(oarch$gh_repo, gh_commit),
+                gh_commits = map(oarch$gh_repo, gh_commit)
 )
@@
   ggplot(aes(date, cum_commits,
-             fill = fct_reorder(rough_category, cat_total))) +
+             fill = fct_reorder(rough_category, cat_total),
+             colour = fct_reorder(rough_category, cat_total))) +
   geom_area() +
   scale_x_datetime(limits = c(ymd("2008-01-01", tz = "UTC"), NA),
-                   date_breaks = "2 years",
+                   date_breaks = "1 years",
                    date_labels = "%Y") +
-  scale_fill_discrete(guide = guide_legend(reverse = TRUE)) +
+  scale_colour_contrast(guide = NULL) +
+  scale_fill_contrast(guide = guide_legend(reverse = TRUE)) +
   labs(
-    title = "Open source archaeology",
+    title = "Open Archaeology",
     subtitle = "Cumulative commits in GitHub repositories",
     caption = "Data: <https://open-archaeo.info> and the GitHub API",
-    x = NULL, y = "Commits", fill = NULL
+    x = NULL, y = NULL, fill = NULL
   ) +
-  theme(legend.position = "bottom")
+  theme_minimal() +
+  theme(
+    text = element_text(family = "IBM Plex Sans"),
+    legend.position = "bottom", 
+    legend.justification = "right",
+    legend.margin = margin(),
+    legend.box.margin = margin()
+  )
 ```

### 2021-03-12 analysis/paper.Rmd
Add quick analysis of issue usage

@@
 
+```{r plot-gh_issues-histogram}
+oarch %>% 
+  unnest(gh_issues) %>% 
+  drop_na(title) %>% 
+  group_by(repo) %>% 
+  summarise(n = n(), .groups = "drop") %>%
+  ggplot(aes(x = n)) +
+  geom_histogram(binwidth = 1) +
+  labs(
+    title = "Repository issues",
+    caption = "Data: <https://open-archaeo.info> and the GitHub API",
+    x = "Number of issues", y = "Number of repositories"
+  )
+```
+
 ### Repository contributors

### 2021-05-11 analysis/paper.Rmd
Preliminary notes for background section.
Also updated the bib file, which now contains the sources mentioned in the notes.

@@
 ```
+
+## Background
+
+- Lots of reviews out there on sets of tools geared towards specific kinds of applications
+  - i.e. for aerial photo recognition, for photogrammetry, for tablet data collection, etc
+
+- A few papers encouraging use of open source tools and platforms
+  - @schmidt2020, @ducke2012, @orengo2015, @ducke2015, @ross2015, @wells2015
+
+- @jeffrey2012, @beck2012, @bevan2012 on the new kinds of engagements and commitments that openness entails
+  - @hansen2019 on the structural/systemtic challenges involved in maintaining a proper archaeological database
+  - @parsons2013, @kansa2014a on carrying over existing ideas and modes of production
+
+- On community-building
+  - @carman2017 on the community-building and @carman2006 on the social practice of digging; we might extend this to discuss extended network of archaeological practices, which now includes software development
+  - @morgan2012, @richardson2017b, @huggett2004 on encouraging wider participation within open collaboration networks
+  - @morgan2015 and @caraher2019 on anarchic ethos in archaeology, see also responses by @richardson2017; we might draw parallels with similar worldviews in open source software development (i.e.
+hacker and DIY culture)
+  - @sholler2019 wrote a general guide for encouraging participation within open source research projects (but not as an archaeologist)
+
+- @daems2020 on educating students to use digital tools, and @morgan2018a on problems with making tools easy to use (relates to @caraher2015's value of friction)
+  - Might also draw from @graham2019 and @graham2020 on pedagogical and exploratory aspects of hacking away
 

### 2021-06-03 analysis/paper.Rmd
Clean repo names before making requests to GitHub API

@@
 read_csv("https://raw.githubusercontent.com/zackbatist/open-archaeo/master/open-archaeo.csv") %>%
-  mutate(category = na_if(category, "?")) %>%
-  mutate(gh_repo = str_remove(github, "https://github.com/")) ->
+  mutate(category = na_if(category, "?")) ->
   oarch
 
+# Extract clean repos from GitHub URLs
+oarch %>% 
+  mutate(
+    gh_repo = str_match(oarch$github, "^https://github.com/([\\w-\\.]+/[\\w-\\.]+)/?.*?$")[,2]
+  ) ->
+  oarch
+
+# Amalgamate categories into fewer rough categories
 oarch <- mutate(oarch, rough_category = recode(category,

### 2021-06-08 analysis/paper.Rmd
Update analysis

@@
 ```
+
+## Introduction
+
+Research objectives and questions:
+
+* What types of open source projects are being developed by archaeologists?
+  * How many projects are there?
+  * What do they do?
+  * What languages are they written in and/or what platform do they use?
+  * How has this changed over time?
+* To what extent to these projects leverage tools for collaborative software engineering?
+  * How many entries use version control?
+  * How many entries have an explicit license?
+  * What are the most popular licenses?
+  * What are the most popular repository hosts?
+  * How active is the average repository?
+  * How fast have repositories been growing?
+  * How many people contribute to individual repositories?
+  * What roles do contributors to repositories have?
+  * Are networks of collaboration emerging?
+  * Contributions over time: is open source software in archaeology growing?
+  * How frequently used are forks, pull requests, issues, and comments used?
+  * How frequently used are 'engagement' features like stars, follows, and releases?
+  * Collaboration over time: is the use of these features becoming more common?
+* Does collaboration in software development mirror, or differ from, collaborative practices in archaeological research more broadly?
 
@@
 # Amalgamate categories into fewer rough categories
+# TODO: redo
 oarch <- mutate(oarch, rough_category = recode(category,
@@
 
-### Tags and categories
+### Open archaeology
+
+TODO [ZB]: Can you explain the logic behind the taxonomies, either here or in materials?
+
+#### Categories and tags
 
@@
   labs(
-    title = "Open archaeology categories",
-    caption = "Data: <https://open-archaeo.info>",
+    title = "Categories",
+    caption = "Data: open-archaeo.info",
     x = NULL, y = "Number of projects"
@@
 
-```{r plot-tags}
+```{r plot-tags, fig.asp=2}
 oarch %>%
@@
   labs(
-    title = "Open archaeology tags",
-    caption = "Data: <https://open-archaeo.info>",
+    title = "Tags",
+    caption = "Data: open-archaeo.info",
     x = NULL, y = "Number of projects"
@@
 
-### Use of GitHub
+#### Languages and platforms
+
+The GitHub API provides data on the languages used in a repository based on an automated machine learning algorithm.
+Since most repositories use more than one language—if only one for the actual software, and another (e.g.
+HTML) for documentation—this data is presented as a table of languages and the number of bytes of code in that language detected in the repository.
+We can therefore measure the popularity of languages across repositories in a couple of ways: based on how many repositories use a language at all; or based on how many bytes of that language are present.
+
+```{r plot-gh_langs, fig.show="hold", out.width="50%"}
+oarch %>%
+  unnest(gh_langs) %>%
+  drop_na(lang, bytes) %>%
+  group_by(lang) %>%
+  summarise(n = n(), .groups = "drop") %>%
+  arrange(-n) %>%
+  slice(1:10) %>%
+  ggplot(aes(fct_reorder(lang, n, .desc = TRUE), n)) +
+  geom_col() +
+  labs(title = "Repository languages",
+       subtitle = "Number of repositories using top 10 languages",
+       caption = "Data: open-archaeo.info and GitHub API",
+       x = NULL, y = "Number of repositories")
+
+oarch %>%
+  unnest(gh_langs) %>%
+  drop_na(lang, bytes) %>%
+  group_by(lang) %>%
+  summarise(bytes = sum(bytes), .groups = "drop") %>%
+  arrange(-bytes) %>%
+  slice(1:10) %>%
+  ggplot(aes(fct_reorder(lang, bytes, .desc = TRUE), bytes)) +
+  geom_col() +
+  labs(title = "Repository languages",
+       subtitle = "Total bytes of code in top 10 languages",
+       caption = "Data: open-archaeo.info and GitHub API",
+       x = NULL, y = "Bytes")
+```
+
+Looking at number of bytes puts C++ and the web languages (HTML, PHP, and Javascript) on top, but this is probably affected by the relative verboseness of these languages and the popularity of web-based documentation across implementation languages.
+The number of repositories is probably more reliable and shows the popularity of scientific programming languages—R and Python—followed by web languages and then compiled C and C++.
+
+We also have a 'platform' taxonomy, which is a manually-classified description of the broad framework that the individual projects rely on.
+
+```{r plot-platforms}
+oarch %>%
+  group_by(platform) %>%
+  drop_na(platform) %>%
+  summarise(n = n()) %>%
+  ggplot(aes(fct_reorder(platform, n), n)) +
+  geom_col() +
+  coord_flip() +
+  labs(
+    title = "Repository platforms",
+    caption = "Data: open-archaeo.info",
+    x = NULL, y = "Number of projects"
+  )
+```
+
+Here again we see R is overwhelmingly dominant; closely followed by Python.
+Plugins for QGIS and mobile apps are also relatively popular.
+The remainder show a rather fragmented landscape of plugins for other desktop software (e.g. AutoCAD, ArcGIS); lesser used programming languages; and custom form and spreadsheet templates.
+
+### Open source
+
+#### Code availability and license
+
+All entries on open-archaeo include source code in some form, but how many have a version-controlled repository? What repository hosts are the most popular?
+
+```{r plot-repo-hosts}
+# TODO: add "none"
+oarch %>% 
+  pivot_longer(c(github, gitlab, bitbucket, launchpad, codeberg),
+               names_to = "repo_host", values_to = "repo_url") %>% 
+  drop_na(repo_url) %>% 
+  count(repo_host) %>% 
+  arrange(-n) %>%
+  ggplot(aes(fct_reorder(repo_host, n, .desc = TRUE), n)) +
+  geom_col() +
+  labs(title = "Repository hosts",
+       caption = "Data: open-archaeo.info",
+       x = NULL, y = "Number of projects")
+```
+
+How many are explicitly licensed? What are the most popular licenses?
+
+#### Repository activity
+
+(On GitHub:)
+
+  * How active is the average repository?
+  * How fast have repositories been growing?
 
@@
 
-```{r plot-gh_langs, fig.show="hold", out.width="50%"}
+#### Growth
+
+### Collaboration
+
+#### Number of collaborators and roles
+
+```{r plot-gh_contributors-histogram}
 oarch %>%
-  unnest(gh_langs) %>%
-  drop_na(lang, bytes) %>%
-  group_by(lang) %>%
+  drop_na(gh_repo) %>%
+  unnest(gh_contribs) %>%
+  group_by(gh_repo) %>%
   summarise(n = n(), .groups = "drop") %>%
-  arrange(-n) %>%
-  slice(1:10) %>%
-  ggplot(aes(fct_reorder(lang, n, .desc = TRUE), n)) +
-  geom_col() +
-  labs(title = "Languages used in GitHub repositories",
-       subtitle = "Number of repositories using top 10 languages",
+  ggplot(aes(x = n)) +
+  geom_histogram(binwidth = 1) +
+  labs(
+    title = "Repository contributors",
     caption = "Data: <https://open-archaeo.info> and the GitHub API",
-       x = NULL, y = "Number of repositories")
+    x = "Number of contributors", y = "Number of repositories"
+  )
+```
 
-oarch %>%
-  unnest(gh_langs) %>%
-  drop_na(lang, bytes) %>%
-  group_by(lang) %>%
-  summarise(bytes = sum(bytes), .groups = "drop") %>%
-  arrange(-bytes) %>%
-  slice(1:10) %>%
-  ggplot(aes(fct_reorder(lang, bytes, .desc = TRUE), bytes)) +
-  geom_col() +
-  labs(title = "Languages used in GitHub repositories",
-       subtitle = "Total bytes of code in top 10 languages",
-       caption = "Data: <https://open-archaeo.info> and the GitHub API",
-       x = NULL, y = "Bytes")
-```
+#### Forks and pull requests
+
+#### Issues and comments
 
@@
 
-### Repository contributors
+#### Collaborative networks
 
-```{r plot-gh_contributors-histogram}
-oarch %>%
-  drop_na(gh_repo) %>%
-  unnest(gh_contribs) %>%
-  group_by(gh_repo) %>%
-  summarise(n = n(), .groups = "drop") %>%
-  ggplot(aes(x = n)) +
-  geom_histogram(binwidth = 1) +
-  labs(
-    title = "Repository contributors",
-    caption = "Data: <https://open-archaeo.info> and the GitHub API",
-    x = "Number of contributors", y = "Number of repositories"
-  )
-```
+If we construct a network linking repositories by their common contributors, can we identify emerging networks of collaboration, or are individual projects still largely isolated?
+If there are networks, what characterises their links?

### 2021-06-08 analysis/paper.Rmd
Added more text, specifically to intro and materials.
Also added some notes for potentially useful analyses.

@@
 
+ZB note: The following relates to what we were discussing last week, and hopefully is a start at making our research questions more relevant to general discourse.
+We can heavily modify this if we do not want to get too far into the general discourse.
+
+This survey will help identify factors that contribute the simultaneous growth and stagnation of archaeological software development.
+The extreme optimism expressed during the early 2010s regarding the potential value that open source development models would bring (particularly in the 2012 special issue of World Archaeology) has been both realized and slow to take off.
+Open source is either a perpetual novelty or a normalized attitude in different circles.
+In this sense, our study is primarily about the development of norms and practices, and the delegation or distribution of various kinds of activities, throughout the archaeological community.
+
+We do this is by leveraging metrics of contributions made by involved participants.
+These metrics reflect how work is _actually_ delegated or distributed, which tempers and perhaps even contradicts common conceptions about how collaborative software development operates.
+
+We emphasize how on highlighting the unique flavour of open source software development that pertains to archaeology specifically.
+It's ok if we are not purists, or if we do not operate in the same ways as other scientific disciplines or Silicon Valley startups.
+We have our own constraints to deal with, and our own distinct goals, mindsets and modes of organizing ourselves that differ from those communities.
+Our goal in this paper is to identify what's going on, to highlight our unique archaeological open source culture, and how that culture has emerged and continues to develop.
+
 ## Background
@@
 We used *open-archaeo* <https://open-archaeo.info/>, a list of open source archaeological software and other digital resources.
-The criteria for inclusion in open-archaeo are [...].
-The search strategy for compiling it was [...].
+As of June 8, 2021, 401 items are listed.
+Items are included in open-archaeo if a project is identified by its maintainers as relating to archaeological work.
+
+The dataset was compiled by browsing collaborative software development platforms, essentially manually crawling through the profiles of users who identify as archaeologists or who have contributed to tools that pertain to archaeological work.
+This quasi-systematic collection strategy was also supplemented by contributions made by interested individuals who identified relevant work that was initially overlooked.
+
+Open-archaeo is a relatively comprehensive list, however it generally lacks code written before the growth and general use of collaborative software development platforms among archaeologists (which seems to have began during the early-2010s) and code that is not available on the web.
+Additionally, items pertaining to various specialist domains may be spotty, owing to the limited experiences of the list's primary maintainers.
+To help resolve this issue, open-archaeo has put out a call specialists to perform 'deep dives' into the software ecosystems pertaining to their methodological or subject domains.
+Thus far, one 'deep dive' into tools used in archaeogenetics has been initiated.
 
@@
 ```
+
+Each record is also annotated with tags that describe the aspects of archaeological work that each tool contributes to.
+Items are tagged based on how they have been identified or presented by their maintainers.
+The list of tags and their associated definitions is subject to grow or change as archaeologists continue to develop software for additional use cases.
+
+Records are also categorized based on how each tool or resource is accessed or used.
+This gives us a sense of the pervasiveness of various development models, and the requirements or dependent knowledge that devs impose on their users.
+
+Two kinds of categories have been defined: software and documents.
+Software comprise collections of instructions that dictate how a computer should operate.
+Documents are bounded sets of recorded information.
+Three categories of software and four categories of documents are defined below.
+
+|Category|Kind|Scope|
+|:-------|:---|:----|
+|Packages and libraries|Software|Sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform.|
+|Standalone software|Software|Software that may be operated without needing to first access an underlying platform.|
+|Scripts|Software|Sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation.|
+|Specifications, protocols and schemas|Documents|A formal data structure or framework intended to be used as a model.|
+|Products|Documents|Stable outcomes of creative work.|
+|Guides|Documents|An educational resource or documented protocol meant to instruct readers how to apply relevant tools or techniques.|
+|Lists and datasets|Documents|A series of consistently organized observations assembled with purpose.|
 
@@
 ```
+ZB note: How many users per platform? Codeberg is second but only has one user.
 
@@
 ```
+ZB note / idea: Plot the dates of the first commits from each project to track the initiation of projects over time.
+Then a secondary analysis does the same thing but with a filter to limit the dataset to include only those projects with a certain minimum number of commits or issues (or whatever metric for activity we plan to use), or that have had activity during the past 3 months, i.e.
+what are the ages of projects that actually had active or ongoing development, or that are still active today?
+This might indicate the average lifespans of projects, or the properties of particularly active projects can be dissected to identify common practices that we can infer are the causes of their liveliness.
 
@@
 If there are networks, what characterises their links?
+
+ZB: I'm speculating that there are going to be various users who have their fingers in many projects, and that those projects tend to be the most active.
+If this is the case, I would further speculate that these people hold a different attitude regarding authorship and the dividuality of projects that is more in sync with the collaborative capabilities of open source, which are at odds with traditional academic norms regarding how contributions are credited and how products are produced are disseminated.

### 2021-06-13 analysis/paper.Rmd
Add some tasks to do

@@
 
+TODO: colour by platform
+
 The GitHub API provides data on the languages used in a repository based on an automated machine learning algorithm.
@@
 ```{r plot-gh_langs, fig.show="hold", out.width="50%"}
+# TODO: some merging of tags
+# TODO: a crosstab/heatmap of tag combinations
 oarch %>%
@@
 # TODO: add "none"
+# TODO: version with unique authors
 oarch %>% 
@@
 
+TODO: barcode plot (see above)
+
 #### Growth
@@
 ### Collaboration
+
+TODO: For this we need **covariaties**! Could be:
+
+* Domain (tags and categories)
+* Time
+* Social aspects: e.g. centrality extracted from network analysis
 

### 2021-06-13 analysis/paper.Rmd
Add tag co-occurrence analysis

@@
 
+Which tags co-occur more often?
+
+```{r plot-tags-combo}
+# NB. Plot in a big window if you want to actually read this!
+oarch %>%
+  pivot_longer(c(tag1, tag2, tag3, tag4, tag5), values_to = "tag",
+               values_drop_na = TRUE) %>%
+  select(item_name, tag) %>% 
+  left_join(., ., by = "item_name") %>% # Join to self
+  group_by(tag.x, tag.y) %>% 
+  count() %>% 
+  filter(tag.x != tag.y) %>% 
+  ggplot(aes(tag.x, tag.y, label = n)) +
+  geom_tile(fill = "black") +
+  geom_text(colour = "white") +
+  scale_fill_davos(guide = guide_none()) +
+  coord_fixed() +
+  labs(title = "Co-occurrence of tags",
+       caption = "Data: open-archaeo.info",
+       x = NULL, y = NULL) +
+  theme(
+    axis.text.x = element_text(angle = 90, hjust = 1)
+  )
+```
+
 #### Languages and platforms

### 2021-06-13 analysis/paper.Rmd
Update to do

@@
 ```{r plot-tags, fig.asp=2}
+# TODO: some merging of tags
 oarch %>%
@@
 ```{r plot-gh_langs, fig.show="hold", out.width="50%"}
-# TODO: some merging of tags
-# TODO: a crosstab/heatmap of tag combinations
 oarch %>%

### 2021-06-13 analysis/paper.Rmd
Add initial project lifespan analysis
@zackbatist Does this do what you suggested?

@@
 ```
+
 ZB note / idea: Plot the dates of the first commits from each project to track the initiation of projects over time.
@@
 This might indicate the average lifespans of projects, or the properties of particularly active projects can be dissected to identify common practices that we can infer are the causes of their liveliness.
+
+```{r plot-lifespan}
+min_commits <- 5
+
+oarch %>% 
+  unnest(gh_commits) %>% 
+  drop_na(datetime) %>% 
+  group_by(item_name) %>%
+  add_tally() %>% 
+  filter(n >= min_commits) %>% 
+  mutate(first_commit = min(datetime),
+         mid_commit = median(datetime),
+         last_commit = max(datetime)) %>%
+  ggplot(aes(x = fct_reorder(item_name, mid_commit, .desc = TRUE),
+             y = datetime,
+             ymin = first_commit,
+             ymax = last_commit)) +
+  geom_linerange() +
+  geom_point(shape = "|") +
+  scale_y_datetime(date_breaks = "1 year") +
+  coord_flip() +
+  labs(
+    title = "Project lifespan",
+    subtitle = glue("Individual commits in GitHub repositories (with {min_commits} or more commits)"),
+    x = NULL, y = NULL,
+    caption = "Data: open-archaeo.info and GitHub API"
+  )
+```
 

### 2021-06-13 analysis/paper.Rmd
Expand lifespan/activity analysis

@@
 
-TODO: barcode plot (see above)
+```{r lifespan-stats}
+oarch %>% 
+  unnest(gh_commits) %>% 
+  drop_na(datetime) %>% 
+  group_by(item_name) %>%
+  summarise(
+    n = n(),
+    first_commit = min(datetime),
+    last_commit = max(datetime),
+    lifespan = as.integer(last_commit - first_commit),
+    commit_rate = n / lifespan
+  ) ->
+  oarch_lifespan
+```
+
+```{r plot-lifespan-hist}
+oarch_lifespan %>% 
+  ggplot(aes(x = lifespan)) +
+  geom_histogram() +
+  labs(
+    title = "Project lifespan",
+    subtitle = "Time between first and latest commit",
+    x = "Days", y = "Repositories",
+    caption = "Data: open-archaeo.info and GitHub API"
+  )
+```
+```{r plot-rate-hist}
+oarch_lifespan %>% 
+  ggplot(aes(x = commit_rate)) +
+  geom_histogram() +
+  labs(
+    title = "Project activity",
+    subtitle = "Average commit rate",
+    x = "Commits per day", y = "Repositories",
+    caption = "Data: open-archaeo.info and GitHub API"
+  )
+```
+
+What are the characteristics of long-lived and/or active repositories?
+
+```{r plot-lifespan-vs-rate}
+oarch_lifespan %>% 
+  ggplot(aes(x = lifespan, y = commit_rate)) +
+  geom_point() +
+  scale_x_log10() +
+  scale_y_log10() +
+  labs(
+    title = "Project lifespan vs. average commit rate",
+    x = "Lifespan (days)", y = "Commits per day",
+    caption = "Data: open-archaeo.info and GitHub API"
+  )
+```
+
+TODO: ZB: Any ideas for other relevant factors? Lots of ways you could go here...
 

### 2021-06-13 analysis/paper.Rmd
Expand lifespan/activity analysis
Adding summaries by tag and category

@@
 
-(On GitHub:)
-
-  * How active is the average repository?
-  * How fast have repositories been growing?
-
 ```{r plot-gh_commits-cum}
@@
 
-ZB note / idea: Plot the dates of the first commits from each project to track the initiation of projects over time.
-Then a secondary analysis does the same thing but with a filter to limit the dataset to include only those projects with a certain minimum number of commits or issues (or whatever metric for activity we plan to use), or that have had activity during the past 3 months, i.e.
-what are the ages of projects that actually had active or ongoing development, or that are still active today?
-This might indicate the average lifespans of projects, or the properties of particularly active projects can be dissected to identify common practices that we can infer are the causes of their liveliness.
+Use of GitHub has been growing consistently since around 2014, but how is this activity distributed across repositories?
 
@@
 ```
+
+So we can see the "lifespan" and activity varies across repositories.
+But how exactly?
 
@@
 
+```{r plot-lifespan-activity-tags}
+#TODO: think we might need some exclusion rules here
+
+left_join(oarch_lifespan, oarch, by = c("item_name")) %>% 
+  pivot_longer(c(tag1, tag2, tag3, tag4, tag5), values_to = "tag",
+               values_drop_na = TRUE) %>%
+  drop_na(tag) %>% 
+  group_by(tag) %>%
+  summarise(avg_lifespan = mean(lifespan),
+            avg_commit_rate = mean(commit_rate)) ->
+  oarch_lifespan_by_tag
+
+oarch_lifespan_by_tag %>% 
+  arrange(-avg_lifespan) %>% 
+  slice(1:10) %>% 
+  ggplot(aes(x = fct_reorder(tag, avg_lifespan, .desc = TRUE),
+             y = avg_lifespan)) +
+  geom_col() +
+  labs(
+    title = "Project lifespan by tag",
+    subtitle = "Top 10 longest-lived tags",
+    x = NULL, y = "Average lifespan (days)",
+    caption = "Data: open-archaeo.info and GitHub API"
+  )
+
+oarch_lifespan_by_tag %>% 
+  arrange(-avg_commit_rate) %>% 
+  slice(1:10) %>% 
+  ggplot(aes(x = fct_reorder(tag, avg_commit_rate, .desc = TRUE),
+             y = avg_commit_rate)) +
+  geom_col() +
+  labs(
+    title = "Project activity by tag",
+    subtitle = "Top 10 most activity tags",
+    x = NULL, y = "Average activity (commits per day)",
+    caption = "Data: open-archaeo.info and GitHub API"
+  )
+```
+
+```{r plot-lifespan-activity-cats}
+#TODO: Eliminate Infs
+
+left_join(oarch_lifespan, oarch, by = c("item_name")) %>% 
+  drop_na(category) %>% 
+  group_by(category) %>%
+  summarise(avg_lifespan = mean(lifespan),
+            avg_commit_rate = mean(commit_rate)) ->
+  oarch_lifespan_by_cat
+
+oarch_lifespan_by_cat %>% 
+  arrange(-avg_lifespan) %>% 
+  slice(1:10) %>% 
+  ggplot(aes(x = fct_reorder(category, avg_lifespan, .desc = TRUE),
+             y = avg_lifespan)) +
+  geom_col() +
+  labs(
+    title = "Project lifespan by category",
+    subtitle = "Top 10 longest-lived categories",
+    x = NULL, y = "Average lifespan (days)",
+    caption = "Data: open-archaeo.info and GitHub API"
+  )
+
+oarch_lifespan_by_cat %>% 
+  arrange(-avg_commit_rate) %>% 
+  slice(1:10) %>% 
+  ggplot(aes(x = fct_reorder(category, avg_commit_rate, .desc = TRUE),
+             y = avg_commit_rate)) +
+  geom_col() +
+  labs(
+    title = "Project activity by categories",
+    subtitle = "Top 10 most activity categories",
+    x = NULL, y = "Average activity (commits per day)",
+    caption = "Data: open-archaeo.info and GitHub API"
+  )
+```
+
 TODO: ZB: Any ideas for other relevant factors? Lots of ways you could go here...
@@
 
-TODO: For this we need **covariaties**! Could be:
+TODO: For this we need **covariates**! Could be:
 

### 2021-06-13 analysis/paper.Rmd
Add total commits to activity vs. lifespan plot

@@
 oarch_lifespan %>% 
-  ggplot(aes(x = lifespan, y = commit_rate)) +
+  ggplot(aes(x = lifespan, y = commit_rate, size = n)) +
   geom_point() +
@@
     title = "Project lifespan vs. average commit rate",
-    x = "Lifespan (days)", y = "Commits per day",
+    x = "Lifespan (days)", y = "Commits per day", size = "Total commits",
     caption = "Data: open-archaeo.info and GitHub API"

### 2021-06-13 analysis/paper.Rmd
Added a bunch of text, some of which is placeholder or notes.
Also rearranged some sections, including some code blocks.

@@
 
-ZB note: The following relates to what we were discussing last week, and hopefully is a start at making our research questions more relevant to general discourse.
-We can heavily modify this if we do not want to get too far into the general discourse.
+Over the past 10-15 years, archaeologists have embraced, at least in theory, the advent of open source software development.
+There has been a lot of optimism regarding the potential value that archaeological software development would bring (particularly in the 2012 special issue of World Archaeology), however depending on who you talk to, creative use of computers among archaeologists has either been a perpetual novelty or a normalized way of working.
 
-This survey will help identify factors that contribute the simultaneous growth and stagnation of archaeological software development.
-The extreme optimism expressed during the early 2010s regarding the potential value that open source development models would bring (particularly in the 2012 special issue of World Archaeology) has been both realized and slow to take off.
-Open source is either a perpetual novelty or a normalized attitude in different circles.
-In this sense, our study is primarily about the development of norms and practices, and the delegation or distribution of various kinds of activities, throughout the archaeological community.
+In this paper, we survey the field of archaeological software development in order to understand how archaeologists are _actually_ creating code.
+We rely rely on a data set called open-archaeo, which is a comprehensive list of software and resources developed by archaeologists for archaeological purposes.
+This is supplemented by data collected by GitHub, a collaborative open source software development platform that hosts the vast majority of projects listed in open-archaeo, and which automatically tracks users' contributions to various projects.
 
-We do this is by leveraging metrics of contributions made by involved participants.
-These metrics reflect how work is _actually_ delegated or distributed, which tempers and perhaps even contradicts common conceptions about how collaborative software development operates.
+The metrics we use reflect how work is distributed across time and delegated among members of a project and among the community as a whole.
+These data are analyzed to highlight how collaborative software development operates within our community.
 
-We emphasize how on highlighting the unique flavour of open source software development that pertains to archaeology specifically.
-It's ok if we are not purists, or if we do not operate in the same ways as other scientific disciplines or Silicon Valley startups.
-We have our own constraints to deal with, and our own distinct goals, mindsets and modes of organizing ourselves that differ from those communities.
-Our goal in this paper is to identify what's going on, to highlight our unique archaeological open source culture, and how that culture has emerged and continues to develop.
+This aspect of _collaborative_ software development is crucial to our interest in the historical development of software engineering among archaeologists.
+Archaeological software development is often (or nearly always) discussed as _open source_ software development.
+A technical definition of open source software is software whose source code is publicly accessible, and which is free to be used, modified or built upon by anyone and for any purpose.
+This implies, but does not necessarily require or benefit from, various collaborative endeavours.
+For instance, code written by a single person that is distributed under an open license, but which is never accessed or reused at any point after it is made available, meets this technical definition of open source, but misses out on the potential value that may be contributed by others.
+A revised definition of open source software defines it as software developed in a collaborative manner, as code that belongs to the community of developers, users and stakeholders who find value in the product and strive for its continued development and iterative improvement.
+We do not deny solitary software developers as members of a broader community; however their behaviour as solitary developers may imply the existence of widespread norms that encourage more basic, but perhaps less fulfilling, ways of working.
+
+In this sense, our study is about the development of norms and practices throughout the archaeological software development community.
+We have our own unique flavour, based on our distinct goals, constraints, mindsets and modes of organizing ourselves as archaeologists, which shape how we write code.
+Our goal in this paper is to highlight how our archaeological open source culture has emerged and continues to develop.
 
@@
 
-#### Categories and tags
+#### Categories, tags, platforms and hosts
+
+So let's start by breaking down the data set.
+
+This plot indicates that most projects included in open-archaeo are designed to be used as a package or library atop a platform.
+The designers of this code are basically creating missing functions within the base platform that are useful for archaeological purposes.
+Some are also creating software to be run independently of any platform.
+A significant amount of projects also comprise of datasets and non-packaged code snippets that have been made available for general use.
 
@@
 ```
+
+```{r plot-platforms}
+oarch %>%
+  group_by(platform) %>%
+  drop_na(platform) %>%
+  summarise(n = n()) %>%
+  ggplot(aes(fct_reorder(platform, n), n)) +
+  geom_col() +
+  coord_flip() +
+  labs(
+    title = "Repository platforms",
+    caption = "Data: open-archaeo.info",
+    x = NULL, y = "Number of projects"
+  )
+```
+
+R is overwhelmingly dominant as a platform, closely followed by Python.
+Plugins for QGIS and mobile apps are also relatively popular.
+The remainder show a rather fragmented landscape of plugins for other desktop software (e.g. AutoCAD, ArcGIS); lesser used programming languages; and custom form and spreadsheet templates.
+
+When we break it down by number of unique contributors, we can account for the popularity of a platform in terms of share of the general population, regardless of how prolific their user-base is.
+
+ZB: Break down platform by contributors, since a few people may be prolific in their support of a platform without necessarily representing the whole population.
+
+If we compare categories with tags, we can see the general domains that each kind of resource is designed to serve.
+
+ZB: Compare tags with categories, to see the purposes of standalone software, datasets, guides, etc.
+
+If we look at tags on their own, we can see which domains of archaeological research software is primarily being developed for.
+However, many items listed in open-archaeo have more than one tag.
+Based on an analysis of the co-occurence of tags, we infer that [...].
 
@@
 ```
-
-Which tags co-occur more often?
 
@@
 ```
+
+The vast majority of projects included in open-archaeo are hosted on GitHub, but a few other hosts are used too.
+If we break this down by number of unique project maintainers, GitHub still remains dominant, but we see that each other platform is only used by a handful of individuals.
+
+```{r plot-repo-hosts}
+# TODO: add "none"
+# TODO: version with unique authors
+oarch %>% 
+  pivot_longer(c(github, gitlab, bitbucket, launchpad, codeberg),
+               names_to = "repo_host", values_to = "repo_url") %>% 
+  drop_na(repo_url) %>% 
+  count(repo_host) %>% 
+  arrange(-n) %>%
+  ggplot(aes(fct_reorder(repo_host, n, .desc = TRUE), n)) +
+  geom_col() +
+  labs(title = "Repository hosts",
+       caption = "Data: open-archaeo.info",
+       x = NULL, y = "Number of projects")
+```
+
+ZB: Plot host by number of unique users.
+
+How many are explicitly licensed? What are the most popular licenses?
+
+We also plotted the development of new software over time.
+
+ZB: Do a temporal analysis to get a sense of when platforms, tags and categories have been gaining popularity.
+There may not be a need for extreme detail here, but simply noting the dates of first commits might be helpful.
+I'm imagining a stacked bar chart, with each bar representing a month or a year, and with the stacked portions of each bar representing a platform or category.
+
+<<< cut? >>>
 
@@
 
-```{r plot-platforms}
-oarch %>%
-  group_by(platform) %>%
-  drop_na(platform) %>%
-  summarise(n = n()) %>%
-  ggplot(aes(fct_reorder(platform, n), n)) +
-  geom_col() +
-  coord_flip() +
-  labs(
-    title = "Repository platforms",
-    caption = "Data: open-archaeo.info",
-    x = NULL, y = "Number of projects"
-  )
-```
-
-Here again we see R is overwhelmingly dominant; closely followed by Python.
-Plugins for QGIS and mobile apps are also relatively popular.
-The remainder show a rather fragmented landscape of plugins for other desktop software (e.g. AutoCAD, ArcGIS); lesser used programming languages; and custom form and spreadsheet templates.
-
-### Open source
-
-#### Code availability and license
-
-All entries on open-archaeo include source code in some form, but how many have a version-controlled repository? What repository hosts are the most popular?
-
-```{r plot-repo-hosts}
-# TODO: add "none"
-# TODO: version with unique authors
-oarch %>% 
-  pivot_longer(c(github, gitlab, bitbucket, launchpad, codeberg),
-               names_to = "repo_host", values_to = "repo_url") %>% 
-  drop_na(repo_url) %>% 
-  count(repo_host) %>% 
-  arrange(-n) %>%
-  ggplot(aes(fct_reorder(repo_host, n, .desc = TRUE), n)) +
-  geom_col() +
-  labs(title = "Repository hosts",
-       caption = "Data: open-archaeo.info",
-       x = NULL, y = "Number of projects")
-```
-ZB note: How many users per platform? Codeberg is second but only has one user.
-
-How many are explicitly licensed? What are the most popular licenses?
+<<< /cut? >>>
 
 #### Repository activity
+
+The next set of analyses relies on the GitHub API to assess the leverage data about the activity occurring within each repository.
+To simplify our analysis and maintain a consistent dataset for the subsequent set of analyses, we excluded data from other hosts such as Codeberg or BitBucket, which cumulativey amount to only n items (XX% of the total).
 
@@
 Use of GitHub has been growing consistently since around 2014, but how is this activity distributed across repositories?
+
+We found that the mean/median/mode lifespan of a project is XX months. Projects that have had some activity during the past 3 months have ages ranging from XX-YY months ago.
+
+ZB: Add columns to the right of the project lifespan plot with values showing commits per day, number of unique contributors, number of days between first and last commit, and any other relevant variables.
+Then discuss relationships between these variables.
 
@@
 ```
+
 ```{r plot-rate-hist}
@@
 
+Projects with a short span of activity tend to have very high commit frequency, whereas there is much more variety in long-term projects.
+Among projects with a greater lifespan, those with more total commits have much more frequent activity than those with fewer total commits, which represent the majority of long-lived projects.
+
+Some of this can be explained in terms of natural processes (i.e.
+the only way to fit a high number of commits within the same span as a project with a low number of commits is to commit more frequently)
+
+But when we break it down by number of contributors [...]
+
 ```{r plot-lifespan-vs-rate}
@@
 ```
+
+The measures of activity may also be compared with the tags.
+
+ZB: Lifespan by tag is dominated by ancient DNA tags.
+Another hint that it may be best to lump all of those.
+In general though, I'm not sure how helpful the following tables are, unless we account for volume somehow.
 
@@
 
+ZB: Add a chart, limited to repos with more than 1 contributor.
+It should present how evenly distributed commits are across all contributors.
+So maybe a percent stacked bar chart, with bars representing a repository, with each bar broken down into sections representing each inidividual contributor, and each section's size representing the percentage of the individual's commits out of the total from the repo.
+
 #### Forks and pull requests
@@
 
+ZB: How could we compare number of commits with number of issues? Some sort of coefficient, like commits divided by issues? Would this even be valuable at all?
+
 #### Collaborative networks

### 2021-06-14 analysis/paper.Rmd
Updating the intro once again.
Hopefully it is more generous and charismatic that my previous rants.

@@
 
-Over the past 10-15 years, archaeologists have embraced, at least in theory, the advent of open source software development.
-There has been a lot of optimism regarding the potential value that archaeological software development would bring (particularly in the 2012 special issue of World Archaeology), however depending on who you talk to, creative use of computers among archaeologists has either been a perpetual novelty or a normalized way of working.
+In applying either physical or digital methods, archaeology has a long tradition of adopting common or 'off the shelf' tools, and leveraging them in ways to serve our own unique needs that are typically unanticipated by their original designers.
 
-In this paper, we survey the field of archaeological software development in order to understand how archaeologists are _actually_ creating code.
-We rely rely on a data set called open-archaeo, which is a comprehensive list of software and resources developed by archaeologists for archaeological purposes.
-This is supplemented by data collected by GitHub, a collaborative open source software development platform that hosts the vast majority of projects listed in open-archaeo, and which automatically tracks users' contributions to various projects.
+A lot of this is surely due to resourcing and budgeting constraints, but holding prerequisite knowledge to create new tools, specifically designed for archaeology, may be a significant factor as well.
 
-The metrics we use reflect how work is distributed across time and delegated among members of a project and among the community as a whole.
-These data are analyzed to highlight how collaborative software development operates within our community.
+And while most archaeologists are probably content to outsource the metallurgical skills needed to forge their trowels, an increasingly large number of us are able to code our own programs and libraries, which serve functions that warrant archaeological input and initiative.
 
-This aspect of _collaborative_ software development is crucial to our interest in the historical development of software engineering among archaeologists.
-Archaeological software development is often (or nearly always) discussed as _open source_ software development.
-A technical definition of open source software is software whose source code is publicly accessible, and which is free to be used, modified or built upon by anyone and for any purpose.
-This implies, but does not necessarily require or benefit from, various collaborative endeavours.
-For instance, code written by a single person that is distributed under an open license, but which is never accessed or reused at any point after it is made available, meets this technical definition of open source, but misses out on the potential value that may be contributed by others.
-A revised definition of open source software defines it as software developed in a collaborative manner, as code that belongs to the community of developers, users and stakeholders who find value in the product and strive for its continued development and iterative improvement.
-We do not deny solitary software developers as members of a broader community; however their behaviour as solitary developers may imply the existence of widespread norms that encourage more basic, but perhaps less fulfilling, ways of working.
+In fact, archaeologists are prolific programmers, as this paper will demonstrate.
 
-In this sense, our study is about the development of norms and practices throughout the archaeological software development community.
-We have our own unique flavour, based on our distinct goals, constraints, mindsets and modes of organizing ourselves as archaeologists, which shape how we write code.
-Our goal in this paper is to highlight how our archaeological open source culture has emerged and continues to develop.
+Here, we present a survey of archaeological software development with a couple goals in mind.
+
+First, we want to see what kinds of software archaeologists are making, and for what general purposes.
+
+And second, we are interested in seeing how archaeologists collaborate in their programming work, and experience open source development processes, which have their own benefits, challenges and organizing principles that may conflict with or relate to means of collaboration and sharing that archaeologists are already generally accustomed to.
+
+In this sense, we want to see how archaeological software is benefitting from, or missing out on, the affordances that open source development models -- particularly, the value added by a broader community of invested stakeholders, the pragmatic implementation of iterative updates, and increased transparency -- and consider how our own unique software development culture emerged and continues to grow.
 

### 2021-06-14 analysis/paper.Rmd
Enhance platforms plot
Now broken down by first author

--- 2021-06-13 analysis/paper.Rmd
@@
 oarch %>%
+  drop_na(category) %>%
   group_by(category) %>%
-  drop_na(category) %>%
   summarise(n = n()) %>%
@@
   labs(
-    title = "Categories",
+    title = "Most popular categories",
+    subtitle = "By number of projects",
     caption = "Data: open-archaeo.info",
-    x = NULL, y = "Number of projects"
+    x = NULL, y = NULL
   )
@@
 oarch %>%
-  group_by(platform) %>%
   drop_na(platform) %>%
-  summarise(n = n()) %>%
-  ggplot(aes(fct_reorder(platform, n), n)) +
-  geom_col() +
+  group_by(platform, author1_name) %>%
+  summarise(n = n(), .groups = "drop_last") %>%
+  mutate(total = sum(n)) %>% 
+  ggplot(aes(fct_reorder(platform, total), n, group = author1_name)) +
+  geom_col(colour = "white") +
   coord_flip() +
   labs(
-    title = "Repository platforms",
+    title = "Most popular platforms",
+    subtitle = "By number of projects, grouped by primary author",
     caption = "Data: open-archaeo.info",
-    x = NULL, y = "Number of projects"
+    x = NULL, y = NULL
   )
@@
 When we break it down by number of unique contributors, we can account for the popularity of a platform in terms of share of the general population, regardless of how prolific their user-base is.
-
-ZB: Break down platform by contributors, since a few people may be prolific in their support of a platform without necessarily representing the whole population.
 

### 2021-06-14 analysis/paper.Rmd
Expand repo usage/host analysis
Also revise the language analysis text, to reflect that new platform analysis

@@
 library("khroma")
+library("gt")
 ```
@@
 
-#### Categories, tags, platforms and hosts
+#### Categories, tags, and platforms
 
@@
 
-The vast majority of projects included in open-archaeo are hosted on GitHub, but a few other hosts are used too.
-If we break this down by number of unique project maintainers, GitHub still remains dominant, but we see that each other platform is only used by a handful of individuals.
-
-```{r plot-repo-hosts}
-# TODO: add "none"
-# TODO: version with unique authors
-oarch %>% 
-  pivot_longer(c(github, gitlab, bitbucket, launchpad, codeberg),
-               names_to = "repo_host", values_to = "repo_url") %>% 
-  drop_na(repo_url) %>% 
-  count(repo_host) %>% 
-  arrange(-n) %>%
-  ggplot(aes(fct_reorder(repo_host, n, .desc = TRUE), n)) +
-  geom_col() +
-  labs(title = "Repository hosts",
-       caption = "Data: open-archaeo.info",
-       x = NULL, y = "Number of projects")
-```
-
-ZB: Plot host by number of unique users.
-
-How many are explicitly licensed? What are the most popular licenses?
-
 We also plotted the development of new software over time.
@@
 
-<<< cut? >>>
-
 #### Languages and platforms
 
-TODO: colour by platform
+TODO: Do we need to use this data, given it says much the same as the 'platform' taxonomy?
 
@@
 
-We also have a 'platform' taxonomy, which is a manually-classified description of the broad framework that the individual projects rely on.
+This largely confirms the pattern seen in our manual classification of the 'platform' of each project.
 
-<<< /cut? >>>
+#### Version control
+
+All projects included in open-archaeo make their source code available in some form, by definition.
+But not all make use of version control.
+
+```{r plot-repo-hosts}
+oarch <- mutate(oarch, none = ifelse(is.na(github) & 
+                                       is.na(gitlab) & 
+                                       is.na(bitbucket) & 
+                                       is.na(launchpad) & 
+                                       is.na(codeberg),
+                                     "", NA))
+
+oarch %>% 
+  pivot_longer(c(github, gitlab, bitbucket, launchpad, codeberg, none),
+               names_to = "repo_host", values_to = "repo_url") %>% 
+  drop_na(repo_url) %>% 
+  group_by(repo_host, author1_name) %>% 
+  summarise(n = n(), .groups = "drop_last") %>% 
+  mutate(total = sum(n)) %>% 
+  ungroup() %>% 
+  mutate(repo_host = fct_reorder(repo_host, total),
+         repo_host = fct_relevel(repo_host, "none")) %>% 
+  ggplot(aes(repo_host, n, group = author1_name)) +
+  geom_col(colour = "#333333") +
+  coord_flip() +
+  labs(title = "Most popular repository hosts",
+       subtitle = "By number of projects, grouped by first author",
+       caption = "Data: open-archaeo.info",
+       x = NULL, y = NULL)
+```
+
+Amongst those that do use version control, the vast majority are hosted on GitHub, but a few other hosts are used too.
+Codeberg is a distant second-place in terms of popularity, but this is largely driven by a single active author.
+
+```{r table-no-repo}
+filter(oarch, none == "") %>% 
+  count(category)
+```
+
+About one in eight projects do not use one of these repository host.
+Some may use a different host, or not make their repositories publicly available, but it is probably safe to assume that the majority simply do not use version control.
+Many of these are standalone documents or code fragments (scripts) hosted, for example, as 'gists', but there are also some pieces of standalone software that apparently do not include version control.
 

### 2021-06-14 analysis/paper.Rmd
Clean up a couple of notes
Updated cumulative commit plot

@@
 ### Open archaeology
-
-TODO [ZB]: Can you explain the logic behind the taxonomies, either here or in materials?
 
@@
 ```{r table-no-repo}
+# TODO: prettify with gt()
 filter(oarch, none == "") %>% 
@@
   mutate(date = floor_date(datetime, "week")) %>%
-  mutate(rough_category = recode(
-    rough_category,
-    "Standalone apps" = "Standalone apps",
-    "Packages (R, Python, etc.)" = "Packages (R, Python, etc.)",
-    .default = "Other")) %>%
-  group_by(rough_category, date) %>%
+  mutate(lumped_category = recode(category,
+                                  "Specifications, protocols and schemas" = "Documents",
+                                  "Products" = "Documents",
+                                  "Guides" = "Documents",
+                                  "Lists and datasets" = "Documents",
+                                  .default = category)) %>% 
+  group_by(lumped_category, date) %>%
   summarise(n_commits = n(), .groups = "drop_last") %>%
@@
   ggplot(aes(date, cum_commits,
-             fill = fct_reorder(rough_category, cat_total),
-             colour = fct_reorder(rough_category, cat_total))) +
+             fill = fct_reorder(lumped_category, cat_total),
+             colour = fct_reorder(lumped_category, cat_total))) +
   geom_area() +
@@
                    date_labels = "%Y") +
-  scale_colour_contrast(guide = NULL) +
-  scale_fill_contrast(guide = guide_legend(reverse = TRUE)) +
+  scale_colour_bright(guide = NULL) +
+  scale_fill_bright(guide = guide_legend(reverse = TRUE)) +
   labs(
-    title = "Open Archaeology",
-    subtitle = "Cumulative commits in GitHub repositories",
+    title = "Growth of open archaeology",
+    subtitle = "Measured by cumulative commits in GitHub repositories",
     caption = "Data: <https://open-archaeo.info> and the GitHub API",
     x = NULL, y = NULL, fill = NULL
-  ) +
-  theme_minimal() +
-  theme(
-    text = element_text(family = "IBM Plex Sans"),
-    legend.position = "bottom",
-    legend.justification = "right",
-    legend.margin = margin(),
-    legend.box.margin = margin()
   )

### 2021-06-14 analysis/paper.Rmd
Tweak plot scales and add a to do

@@
 
-#### Repository activity
+#### Repository lifespan and activity
 
@@
          last_commit = max(datetime)) %>%
-  ggplot(aes(x = fct_reorder(item_name, mid_commit, .desc = TRUE),
+  ggplot(aes(x = fct_reorder(item_name, last_commit, .desc = TRUE),
              y = datetime,
@@
 ```{r plot-lifespan-hist}
+# TODO: add more x breaks
+
 oarch_lifespan %>% 
@@
   ggplot(aes(x = n)) +
-  geom_histogram(binwidth = 1) +
+  geom_histogram() +
+  scale_x_log10() +
   labs(

### 2021-06-15 analysis/paper.Rmd
Begin generating figures for CAA2021 slides

@@
 library("gt")
+```
+
+```{r plot-theme}
+theme_oarch <- function() theme(
+  text = element_text(family = "Lato")
+)
 ```
@@
 ```{r plot-categories}
+oarch <- mutate(oarch, 
+                category = recode(category,
+                                  "Specifications, protocols and schemas" = "Specifications etc.",
+                                  .default = category))
 oarch %>%
@@
     x = NULL, y = NULL
-  )
+  ) +
+  theme_oarch()
+
+# Export for slides (half slide)
+# ggsave("figures/categories.png", width = 127, height = 127, units = "mm")
 ```
@@
   ggplot(aes(fct_reorder(platform, total), n, group = author1_name)) +
-  geom_col(colour = "white") +
+  geom_col(colour = "#333333") +
   coord_flip() +
@@
     title = "Most popular platforms",
-    subtitle = "By number of projects, grouped by primary author",
+    subtitle = "By number of projects, split by primary author",
     caption = "Data: open-archaeo.info",
     x = NULL, y = NULL
-  )
+  ) +
+  theme_oarch()
+
+# Export for slides (half slide)
+# ggsave("figures/platforms.png", width = 127, height = 127, units = "mm")
 ```
@@
   labs(title = "Most popular repository hosts",
-       subtitle = "By number of projects, grouped by first author",
+       subtitle = "By number of projects, split by primary author",
        caption = "Data: open-archaeo.info",
@@
   mutate(lumped_category = recode(category,
-                                  "Specifications, protocols and schemas" = "Documents",
+                                  "Specifications etc." = "Documents",
                                   "Products" = "Documents",
@@
   geom_area() +
-  scale_x_datetime(limits = c(ymd("2008-01-01", tz = "UTC"), NA),
-                   date_breaks = "1 years",
+  scale_x_datetime(limits = c(ymd("2005-01-01", tz = "UTC"), 
+                              ymd("2021-05-31", tz = "UTC")),
+                   date_breaks = "2 years",
                    date_labels = "%Y") +
+  scale_y_continuous(labels = scales::comma) +
   scale_colour_bright(guide = NULL) +
@@
     subtitle = "Measured by cumulative commits in GitHub repositories",
-    caption = "Data: <https://open-archaeo.info> and the GitHub API",
+    caption = "Data: open-archaeo.info and GitHub API",
     x = NULL, y = NULL, fill = NULL
-  )
+  ) +
+  theme_oarch()
+
+# Export for slides (full slide)
+# ggsave("figures/commits-cum.png", width = 254, height = 127, units = "mm")
 ```
@@
 ```{r plot-lifespan}
-min_commits <- 5
-
 oarch %>% 
@@
   add_tally() %>% 
-  filter(n >= min_commits) %>% 
+  filter(n > 1) %>% 
   mutate(first_commit = min(datetime),
@@
          last_commit = max(datetime)) %>%
-  ggplot(aes(x = fct_reorder(item_name, last_commit, .desc = TRUE),
+  ggplot(aes(x = fct_reorder(item_name, last_commit, .desc = FALSE),
              y = datetime,
@@
              ymax = last_commit)) +
-  geom_linerange() +
-  geom_point(shape = "|") +
-  scale_y_datetime(date_breaks = "1 year") +
-  coord_flip() +
+  geom_linerange(size = 0.05) +
+  geom_point(size = 0.1) +
+  scale_y_datetime(limits = c(ymd("2005-01-01", tz = "UTC"), 
+                              ymd("2021-05-31", tz = "UTC")),
+                   date_breaks = "2 years", date_labels = "%Y") +
+  # coord_flip() +
   labs(
     title = "Project lifespan",
-    subtitle = glue("Individual commits in GitHub repositories (with {min_commits} or more commits)"),
+    subtitle = glue("Individual commits in GitHub repositories (with more than 1 commit)"),
     x = NULL, y = NULL,
     caption = "Data: open-archaeo.info and GitHub API"
+  ) +
+  theme_oarch() +
+  theme(
+    axis.line.x = element_blank(),
+    axis.ticks.x = element_blank(),
+    axis.text.x = element_blank(),
+    panel.grid.minor.x = element_blank(),
+    panel.grid.major.x = element_blank()
   )
+
+# Export for slides (full slide)
+# ggsave("figures/lifespan.png", width = 254, height = 127, units = "mm")
 ```
@@
   )
+
 ```

### 2021-06-15 analysis/paper.Rmd
Add tag/category intersection analysis
Plus adjust breaks in project lifespan histogram

@@
 ```{r plot-tags, fig.asp=2}
-# TODO: some merging of tags
+# Tags to be excluded as overly detailed/duplicating others
+detail_tags <- c("Instrumental Neutron activation analysis",
+                 "Harris Matrix",
+                 "aDNA Simulators",
+                 "Damage profilers",
+                 "Metagenomic classifers",
+                 "aDNA simulators",
+                 "aDNA damage-aware genotypers",
+                 "Low-coverage contamination estimation",
+                 "Low-coverage adapted population-genetics tools",
+                 "aDNA preprocessing",
+                 "aDNA aligner",
+                 "aDNA assembler",
+                 "aDNA toolkit")
+
 oarch %>%
@@
                values_drop_na = TRUE) %>%
+  drop_na(tag) %>%
+  filter(!tag %in% detail_tags) %>% 
   group_by(tag) %>% 
-  drop_na(tag) %>%
   summarise(n = n()) %>%
@@
   labs(
-    title = "Tags",
+    title = "Most popular tags",
     caption = "Data: open-archaeo.info",
@@
   )
+```
+
+```{r plot-tags-cats}
+oarch %>%
+  pivot_longer(c(tag1, tag2, tag3, tag4, tag5), values_to = "tag",
+               values_drop_na = TRUE) %>%
+  drop_na(tag) %>%
+  filter(!tag %in% detail_tags) %>% 
+  mutate(category = recode(category,
+                                  "Specifications etc." = "Documents",
+                                  "Products" = "Documents",
+                                  "Guides" = "Documents",
+                                  "Lists and datasets" = "Documents",
+                                  .default = category)) %>% 
+  group_by(tag, category) %>% 
+  summarise(n = n(), .groups = "drop_last") %>%
+  mutate(total = sum(n)) %>% 
+  ggplot(aes(fct_reorder(tag, total), n, fill = category)) +
+  # facet_wrap(vars(category)) +
+  geom_col() +
+  scale_fill_bright() +
+  coord_flip() +
+  labs(
+    title = "Tags by category",
+    caption = "Data: open-archaeo.info",
+    x = NULL, y = "Number of projects",
+    fill = NULL
+  ) +
+  theme_oarch() +
+  theme(
+    axis.text.y = element_text(size = 7)
+  )
+
+# Export for slides (full slide)
+# ggsave("figures/tags-cats.png", width = 254, height = 127, units = "mm")
 ```
@@
        x = NULL, y = NULL)
+
+# Export for slides (half slide)
+# ggsave("figures/repo-hosts.png", width = 127, height = 127, units = "mm")
 ```
@@
   ggplot(aes(x = lifespan)) +
-  geom_histogram() +
+  geom_histogram(binwidth = 100) +
+  scale_x_continuous(breaks = scales::breaks_width(1000)) +
   labs(

### 2021-06-15 analysis/paper.Rmd
Expand lifespan-vs-rate analysis

@@
   group_by(item_name) %>%
+  mutate(n_contribs = map_int(gh_contribs, vctrs::vec_size)) %>% 
   summarise(
+    category = first(category),
+    n_contribs = first(n_contribs),
     n = n(),
@@
 ```{r plot-lifespan-hist}
-# TODO: add more x breaks
-
 oarch_lifespan %>% 
@@
     title = "Project lifespan",
-    subtitle = "Time between first and latest commit",
-    x = "Days", y = "Repositories",
+    subtitle = "Time between first and latest commit, by repository",
+    x = "Days", y = NULL,
+    caption = "Data: open-archaeo.info and GitHub API"
+  ) +
+  theme_oarch()
+
+# Export for slides (half slide)
+# ggsave("figures/lifespan-hist.png", width = 127, height = 127, units = "mm")
+```
+
+```{r plot-lifespan-hist-byyear}
+oarch_lifespan %>% 
+  mutate(first_commit = year(first_commit)) %>% 
+  filter(first_commit %in% 2012:2020) %>% 
+  ggplot(aes(x = lifespan)) +
+  facet_wrap(vars(first_commit), nrow = 1) +
+  geom_histogram(binwidth = 100) +
+  scale_x_continuous(breaks = scales::breaks_width(1000)) +
+  labs(
+    title = NULL,
+    subtitle = NULL,
+    x = NULL, y = NULL,
     caption = "Data: open-archaeo.info and GitHub API"
@@
   ggplot(aes(x = lifespan, y = commit_rate, size = n)) +
-  geom_point() +
+  geom_point(shape = 21, fill = NA) +
   scale_x_log10() +
@@
     x = "Lifespan (days)", y = "Commits per day", size = "Total commits",
+    caption = NULL
+  ) +
+  theme_oarch()
+# Export for slides (half slide)
+# ggsave("figures/lifespan-vs-rate-total.png", width = 127, height = 60, units = "mm")
+
+oarch_lifespan %>% 
+  ggplot(aes(x = lifespan, y = commit_rate, size = n_contribs)) +
+  geom_point(shape = 21, fill = NA) +
+  scale_x_log10() +
+  scale_y_log10() +
+  labs(
+    title = NULL,
+    x = "Lifespan (days)", y = "Commits per day", size = "Contributors",
     caption = "Data: open-archaeo.info and GitHub API"
-  )
+  ) +
+  theme_oarch()
+# Export for slides (half slide)
+# ggsave("figures/lifespan-vs-rate-contrib.png", width = 127, height = 60, units = "mm")
+
 ```
@@
 
-ZB: Lifespan by tag is dominated by ancient DNA tags.
-Another hint that it may be best to lump all of those.
-In general though, I'm not sure how helpful the following tables are, unless we account for volume somehow.
-
 ```{r plot-lifespan-activity-tags}
-#TODO: think we might need some exclusion rules here
-
 left_join(oarch_lifespan, oarch, by = c("item_name")) %>% 

### 2021-06-15 analysis/paper.Rmd
Expand analysis of contributors

@@
     title = "Repository contributors",
-    caption = "Data: <https://open-archaeo.info> and the GitHub API",
-    x = "Number of contributors", y = "Number of repositories"
+    x = "Number of contributors", y = "Repositories",
+    caption = "Data: open-archaeo.info and GitHub API"
+  ) +
+  theme_oarch()
+
+# Export for slides (half slide)
+# ggsave("figures/contrib-hist.png", width = 127, height = 127, units = "mm")
+```
+
+```{r plot-contrib-dist}
+oarch %>% 
+  unnest(gh_contribs) %>% 
+  drop_na(contributor) %>% 
+  group_by(item_name) %>% 
+  mutate(n_contributors = n()) %>% 
+  filter(n_contributors > 1) %>% 
+  mutate(big_boy = ifelse(contributions == max(contributions), 
+                          "Main contributor", "Other contributors")) %>% 
+  ggplot(aes(x = item_name, y = contributions, 
+             group = contributor,
+             fill = big_boy)) +
+  geom_col(position = "fill", colour = "white") +
+  scale_y_continuous(labels = scales::percent) +
+  labs(x = "Repository", y = "Contributions",
+       fill = NULL) +
+  coord_flip() +
+  theme_oarch() +
+  theme(
+    axis.text.y = element_blank(),
+    axis.ticks.y = element_blank()
   )
 
+# Export for slides (half slide)
+# ggsave("figures/contrib-dist.png", width = 127, height = 127, units = "mm")
 ```
-
-ZB: Add a chart, limited to repos with more than 1 contributor.
-It should present how evenly distributed commits are across all contributors.
-So maybe a percent stacked bar chart, with bars representing a repository, with each bar broken down into sections representing each inidividual contributor, and each section's size representing the percentage of the individual's commits out of the total from the repo.
 

### 2021-06-15 analysis/paper.Rmd
Add a bit more basic issues/comments analysis

@@
 
-```{r plot-gh_issues-histogram}
+```{r plot-issues-hist}
 oarch %>%
@@
   ggplot(aes(x = n)) +
-  geom_histogram(binwidth = 1) +
+  geom_histogram() +
+  scale_x_log10() +
   labs(
-    title = "Repository issues",
-    caption = "Data: <https://open-archaeo.info> and the GitHub API",
-    x = "Number of issues", y = "Number of repositories"
-  )
+    title = "Use of issues",
+    caption = "Data: open-archaeo.info and GitHub API",
+    x = "Number of issues", y = "Repositories"
+  ) +
+  theme_oarch()
+
+# Export for slides (half slide)
+# ggsave("figures/issues-hist.png", width = 127, height = 127, units = "mm")
+```
+
+```{r plot-comments-hist}
+oarch %>%
+  unnest(gh_comments) %>%
+  drop_na(issue_num) %>%
+  group_by(repo) %>%
+  summarise(n = n(), .groups = "drop") %>%
+  ggplot(aes(x = n)) +
+  geom_histogram() +
+  scale_x_log10() +
+  labs(
+    title = "Use of comments",
+    caption = "Data: open-archaeo.info and GitHub API",
+    x = "Number of comments", y = "Repositories"
+  ) +
+  theme_oarch()
+
+# Export for slides (half slide)
+# ggsave("figures/comments-hist.png", width = 127, height = 127, units = "mm")
 ```

### 2021-06-16 analysis/paper.Rmd
Revise contrib distribution figure

@@
   mutate(big_boy = ifelse(contributions == max(contributions), 
-                          "Main contributor", "Other contributors")) %>% 
-  ggplot(aes(x = item_name, y = contributions, 
-             group = contributor,
-             fill = big_boy)) +
-  geom_col(position = "fill", colour = "white") +
+                          "Main contributor", "Other contributors"),
+         big_boy_pct = max(contributions) / sum(contributions)) ->x %>% 
+  ggplot(aes(x = fct_reorder(item_name, big_boy_pct, .desc = FALSE),
+             y = contributions, 
+             # group = contributor,
+             fill = fct_relevel(big_boy, c("Other contributors",
+                                           "Main contributor")))) +
+  geom_col(position = "fill", colour = "white", size = 0.1) +
   scale_y_continuous(labels = scales::percent) +
-  labs(x = "Repository", y = "Contributions",
-       fill = NULL) +
+  scale_fill_contrast(guide = guide_legend(reverse = TRUE)) +
+  labs(title = "Repository contributors by proportion of contributions",
+       x = NULL, y = "Contributions",
+       fill = NULL,
+       caption = "Data: open-archaeo.info and GitHub API") +
   coord_flip() +
